---
title: "Task 14 simulation report: lag misspecification and treatment persistence"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  pdf_document:
    toc: true
    number_sections: true
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{float}
params:
  input_dir: !r NULL
---

```{r setup, include=FALSE}
options(scipen = 999)
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.path = file.path(params$input_dir, "figures", "report-")
)
if (is.null(params$input_dir) || !dir.exists(params$input_dir)) {
  stop("params$input_dir must identify a completed Task 14 results directory.")
}
library(data.table)
library(ggplot2)
library(knitr)

input_file <- function(name) file.path(params$input_dir, name)
grid <- fread(input_file("design_grid.csv"))
summary <- fread(input_file("scenario_estimator_summary.csv"))
selection <- fread(input_file("selection_recovery_summary.csv"))
displacement <- fread(input_file("displacement_summary.csv"))
residual_acf <- fread(input_file("residual_autocorrelation_summary.csv"))
validation <- fread(input_file("validation.csv"))
manifest <- fread(input_file("run_manifest.csv"))
failures <- fread(input_file("failures.csv"))
format_values <- function(values) paste(sort(unique(values)), collapse = ", ")

control_labels <- c(
  no_Z = "No Z control",
  Z_lag = "Lagged Z (Z[t-1])",
  Z_contemporaneous = "Contemporaneous Z (Z[t])"
)
summary[, control_label := unname(control_labels[control_spec])]
residual_acf[, control_label := unname(control_labels[control_spec])]
```

# Design and execution record

This execution contains `r nrow(grid)` scenarios: true ADL order `r format_values(grid$true_lag_order)`, treatment persistence `rho_D` equal to `r format_values(grid$rho_D)`, and first-lag direct carryover equal to `r format_values(grid$carryover)`. Every scenario uses $N = `r unique(grid$N)`$, $T = `r unique(grid$T)`$, and the inherited dual-role timing. The execution record reports `r manifest$repetitions[1]` replications per cell and `r manifest$actual_rows[1]` retained raw rows. There are `r nrow(failures)` retained failed fits; they are not removed from any file.

```{r table-design, results='asis'}
design_table <- unique(grid[, .(true_lag_order, rho_D, carryover, N, T)])[order(true_lag_order, rho_D, carryover)]
kable(
  design_table,
  format = "latex",
  booktabs = TRUE,
  position = "H",
  caption = "Pre-specified simulation cells."
)
```

The numerical target is the contemporaneous CET. `Z[t-1]` is the pre-exposure control used by the primary models. `Z[t]` is post-treatment in this DGP and is displayed only to diagnose the specification shift; its numerical coverage is not evidence of causal validity.

# Recovery of the true lag order

AIC and BIC choose from orders one through three using the `Z[t-1]` specifications only. They are never selected ex post. The table averages cell-level recovery rates over the persistence and carryover grid; the individual-cell rates remain in `selection_recovery_summary.csv`.

```{r table-recovery, results='asis'}
recovery_table <- selection[, .(
  mean_recovery_rate = mean(recovery_rate),
  mean_selected_lag_1_rate = mean(selected_lag_1_rate),
  mean_selected_lag_2_rate = mean(selected_lag_2_rate),
  mean_selected_lag_3_rate = mean(selected_lag_3_rate),
  mean_failure_count = mean(n_failed)
), by = .(true_lag_order, selection_criterion)][order(true_lag_order, selection_criterion)]
data.table::setnames(
  recovery_table,
  c(
    "true_lag_order", "selection_criterion", "mean_recovery_rate",
    "mean_selected_lag_1_rate", "mean_selected_lag_2_rate",
    "mean_selected_lag_3_rate", "mean_failure_count"
  ),
  c("True order", "Criterion", "Recovery", "Pr(ADL 1)", "Pr(ADL 2)", "Pr(ADL 3)", "Mean failures")
)
kable(
  recovery_table,
  format = "latex",
  digits = 3,
  booktabs = TRUE,
  position = "H",
  caption = "Lag-selection recovery, reported separately for AIC and BIC."
)
```

```{r fig-recovery, fig.width=8, fig.height=5, fig.pos='H', fig.cap='Correct-lag recovery by information criterion, true lag order, treatment persistence, and carryover.'}
ggplot(selection, aes(x = factor(rho_D), y = factor(carryover), fill = recovery_rate)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", recovery_rate)), size = 3) +
  facet_grid(selection_criterion ~ true_lag_order, labeller = label_both) +
  scale_fill_viridis_c(limits = c(0, 1), name = "Recovery") +
  labs(x = expression(rho[D]), y = "First-lag carryover") +
  theme_minimal(base_size = 10) +
  theme(panel.grid = element_blank())
```

# CET recovery, residual dependence, and specification shifts

Table 3 aggregates the primary `Z[t-1]` specification equally across the saved cells. It reports signed relative bias; a negative value therefore indicates downward rather than absolute bias. Complete scenario-by-scenario results, including no-`Z` and `Z[t]` fits, are retained in `scenario_estimator_summary.csv`.

```{r table-metrics, results='asis'}
metric_table <- summary[control_spec == "Z_lag", .(
  mean_bias = mean(bias),
  mean_relative_bias = mean(relative_bias),
  mean_rmse = mean(rmse),
  mean_coverage = mean(coverage_cet),
  mean_residual_acf1 = mean(mean_residual_acf1)
), by = .(estimator)][order(estimator)]
data.table::setnames(
  metric_table,
  c(
    "estimator", "mean_bias", "mean_relative_bias", "mean_rmse", "mean_coverage",
    "mean_residual_acf1"
  ),
  c("Lag rule", "Bias", "Relative bias", "RMSE", "Coverage", "Residual ACF(1)")
)
kable(
  metric_table,
  format = "latex",
  digits = 3,
  booktabs = TRUE,
  position = "H",
  caption = "CET recovery and residual autocorrelation for the primary Z[t-1] specification."
)
```

```{r fig-bias, fig.width=8, fig.height=8.4, fig.pos='H', fig.cap='Bias of selected primary Z[t-1] specifications across persistence and carryover conditions.'}
primary_bias <- summary[
  control_spec == "Z_lag" & estimator_id %in% c("adl_1", "adl_oracle", "adl_aic", "adl_bic")
]
ggplot(primary_bias, aes(x = factor(rho_D), y = factor(carryover), fill = bias)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.3f", bias)), size = 2.5) +
  facet_grid(estimator ~ true_lag_order, scales = "free_y", labeller = label_both) +
  scale_fill_gradient2(low = "#b2182b", mid = "white", high = "#2166ac", midpoint = 0, name = "Bias") +
  labs(x = expression(rho[D]), y = "First-lag carryover") +
  theme_minimal(base_size = 9) +
  theme(panel.grid = element_blank(), strip.text.y = element_text(size = 8))
```

```{r table-shifts, results='asis'}
shift_table <- displacement[, .(
  mean_delta_lagged_Z = mean(mean_delta_z_lag_minus_none),
  mean_delta_contemporaneous_Z = mean(mean_delta_z_contemporaneous_minus_none)
), by = .(estimator, selection_criterion)][order(estimator)]
data.table::setnames(
  shift_table,
  c("estimator", "selection_criterion", "mean_delta_lagged_Z", "mean_delta_contemporaneous_Z"),
  c("Lag rule", "Selection rule", "Delta: Z[t-1] minus no Z", "Delta: Z[t] minus no Z")
)
kable(
  shift_table,
  format = "latex",
  digits = 3,
  booktabs = TRUE,
  position = "H",
  caption = "Mean specification shifts associated with Z[t-1] and Z[t], relative to no Z control."
)
```

```{r fig-acf, fig.width=8, fig.height=4.5, fig.pos='H', fig.cap='Mean residual lag-one autocorrelation by dynamic and control specification.'}
acf_plot_data <- residual_acf[, .(
  mean_residual_acf1 = mean(mean_residual_acf1)
), by = .(estimator, control_label)]
ggplot(acf_plot_data, aes(x = estimator, y = mean_residual_acf1, fill = control_label)) +
  geom_col(position = position_dodge(width = 0.8)) +
  geom_hline(yintercept = 0, linewidth = 0.35) +
  labs(x = NULL, y = "Mean residual ACF(1)", fill = "Control specification") +
  theme_minimal(base_size = 9) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

# Validation and interpretation

```{r table-validation, results='asis'}
validation_display <- copy(validation)
data.table::setnames(validation_display, c("check", "passed", "detail"), c("Validation check", "Passed", "Detail"))
kable(
  validation_display,
  format = "latex",
  booktabs = TRUE,
  position = "H",
  caption = "Mechanical validation of the completed simulation outputs."
)
```

The results delimit a scope condition, not a mechanical universal rule. An ADL(1) specification is a defensible baseline only when substantive timing supports a one-period history and residual diagnostics do not indicate material omitted dynamics. AIC and BIC are useful, separately reported sensitivity diagnostics for conditional-mean lag order; they cannot determine whether a candidate control is pre-exposure or whether the CET is identified. The `Z[t]` shifts illustrate exactly why contemporaneous controls must not be justified by information criteria alone.

The suggested one-sentence main-text insertion and compact scope-table row are saved in `editorial_recommendation.md`. Full results belong in the appendix; no manuscript file is changed by this package.
