---
title: "Appendix Note: Dual-Role Figure and Firewall Mechanism"
date: "2026-07-11"
output:
  pdf_document:
    latex_engine: pdflatex
    number_sections: true
header-includes:
  - \usepackage{booktabs}
  - \usepackage{float}
  - \usepackage{graphicx}
geometry: margin=1in
fontsize: 10pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(readr)
library(knitr)

artifact_dir <- "."
validation <- readr::read_csv(file.path(artifact_dir, "validation.csv"), show_col_types = FALSE)
manifest <- readr::read_csv(file.path(artifact_dir, "source_manifest.csv"), show_col_types = FALSE)
```

# Scope and inputs

This note documents the reconstruction of the dual-role Figure 2 and the firewall table without running, sourcing, or modifying a simulation. The generation script reads only these existing result artifacts:

```{r source-files}
knitr::kable(manifest, format = "latex", booktabs = TRUE, caption = "Source result artifacts used to generate the Task 11 figure and table.")
```

The central DGP has $Z_{t-1}$ as a confounder of $D_t$ and $Y_t$, while contemporaneous $Z_t$ is a collider of $D_t$ and $Y_t$. The estimand is the contemporaneous treatment effect $\beta=1$. The eight-model result artifact uses $N=100$, $T=30$, a burn-in of 100 periods, and 500 replications in each of its 10 scenarios. Its displayed models are pooled short/long, TWFE short/long, ADL short/long without fixed effects, and ADL short/long with unit and time fixed effects. The Figure 2 subset fixes $\sigma_{\alpha_Z}=0.5$ and displays TWFE short, TWFE plus $Z_{t-1}$, ADL+FE, and ADL+FE plus $Z_{t-1}$.

The firewall artifact varies $(\rho_Z,\rho_Y)$ over 12 scenarios and reports 500 replications per scenario. It compares the long TWFE specification with the long ADL+FE specification. The existing dual-role report identifies this as the same $N=100$, $T=30$ family; the Task 10 inventory records 500 replications but leaves the firewall row's $N$ and $T$ as source metadata. No claim here uses a new draw or an unrecorded simulation run.

# Monte Carlo construction and fidelity check

Bias is the simulated mean of $\widehat\beta-1$; RMSE is $\sqrt{\operatorname{mean}[(\widehat\beta-1)^2]}$. For the eight-model raw file, the Monte Carlo standard error (MCSE) is the sample standard deviation of $\widehat\beta$ across replications divided by $\sqrt{500}$. The plotted 95% Monte Carlo uncertainty interval is mean bias plus or minus $1.96\times\mathrm{MCSE}$. These quantify simulation noise in the estimated mean bias; they are not confidence intervals for an estimator in a single dataset. The firewall CSV supplies its published MCSEs, so the table applies the same formula to those reported quantities.

```{r validation-table}
knitr::kable(validation, format = "latex", booktabs = TRUE, caption = "Reconstruction checks. The figure/table generator stops if recalculated means or MCSEs fail the stated tolerance.")
```

The raw recalculation reproduces the published eight-model bias means and MCSEs to numerical precision before any presentational change. The maximum discrepancy is recorded in the validation artifact rather than silently replaced.

# Reconstructed Figure 2

\begin{figure}[H]
\centering
\includegraphics[width=0.95\linewidth]{figure_2_dual_role_bias_rmse.pdf}
\caption{Dual-role DGP: bias relative to the true contemporaneous treatment effect and RMSE across collider persistence. The DGP has $Z_{t-1}$ as a confounder and $Z_t$ as a contemporaneous collider; the estimand is $\beta=1$. The displayed subset sets $\sigma_{\alpha_Z}=0.5$, with $N=100$, $T=30$, a burn-in of 100 periods, and 500 Monte Carlo replications per scenario. Models are TWFE short, TWFE plus $Z_{t-1}$, ADL+FE, and ADL+FE plus $Z_{t-1}$; FE models include unit and time effects. Error bars are 95\% Monte Carlo uncertainty intervals for simulated mean bias, not estimator confidence intervals.}
\end{figure}

# Firewall mechanism table

\input{table_firewall_mc.tex}

# Interpretation and limitations

For this DGP, conditioning on the lagged outcome and lagged control is associated with a markedly smaller simulated bias than the long TWFE specification when outcome persistence is positive. When $\rho_Y=0$, the long TWFE bias is itself approximately zero, so the contrast is not evidence that the lagged outcome universally improves a specification. It is consistent with the displayed mechanism in which outcome persistence transmits the inherited path; it does not apportion the contrast between d-separation, dynamic conditional-mean specification, or finite-$T$ estimator behavior.

The evidence is conditional on this linear, homogeneous DGP and its stated grid. The firewall result is a published summary CSV rather than a raw-replication artifact, so its MC intervals are reconstructed from reported MCSEs rather than recalculated from individual draws. Neither the intervals nor the figure establish estimator-level sampling coverage.
