---
title: "Finite-T Dynamic-Panel Bias and Estimator Choice"
author: "IVB-paper replication package"
date: "2026-07-11"
output:
  pdf_document:
    latex_engine: xelatex
    number_sections: true
params:
  input_dir: "results"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  out.width = "95%"
)
options(scipen = 999)
required <- c("data.table", "dplyr", "ggplot2", "knitr")
available <- vapply(required, requireNamespace, logical(1), quietly = TRUE)
if (any(!available)) {
  stop("Missing report packages: ", paste(names(available)[!available], collapse = ", "))
}
input_dir <- normalizePath(params$input_dir, mustWork = TRUE)
required_files <- c(
  "design_grid.csv",
  "scenario_estimator_summary.csv",
  "stability_checks.csv",
  "failures.csv",
  "run_manifest.csv"
)
paths <- file.path(input_dir, required_files)
if (any(!file.exists(paths))) {
  stop("Missing full-grid outputs: ", paste(required_files[!file.exists(paths)], collapse = ", "))
}
summary_results <- data.table::fread(file.path(input_dir, "scenario_estimator_summary.csv"))
stability <- data.table::fread(file.path(input_dir, "stability_checks.csv"))
failures <- data.table::fread(file.path(input_dir, "failures.csv"))
run_manifest <- data.table::fread(file.path(input_dir, "run_manifest.csv"))
```

# Pre-specified design

The report reads versioned outputs and does not run simulations. The principal grid contains $N\in\{50,100,250\}$, $T\in\{8,10,15,20,30,50\}$, and $\rho_Y\in\{0.2,0.5,0.8\}$. The stress grid fixes $N=100$ and $\rho_Y=0.5$, varies $T\in\{10,20,30\}$ and $\rho_D\in\{0.2,0.5,0.8\}$, and uses the inherited dual-role unit-effect scales $\sigma_{\alpha Z}\in\{0.5,2.0\}$.

```{r table-design}
design_table <- summary_results |>
  dplyr::distinct(design_family, N, T, rho_Y, rho_D, sigma_alpha_Z) |>
  dplyr::group_by(design_family) |>
  dplyr::summarise(
    scenarios = dplyr::n(),
    N_values = paste(sort(unique(N)), collapse = ", "),
    T_values = paste(sort(unique(T)), collapse = ", "),
    rho_Y_values = paste(sort(unique(rho_Y)), collapse = ", "),
    rho_D_values = paste(sort(unique(rho_D)), collapse = ", "),
    sigma_alpha_Z_values = paste(sort(unique(sigma_alpha_Z)), collapse = ", "),
    .groups = "drop"
  )
knitr::kable(
  design_table,
  col.names = c("Design", "Scenarios", "N", "T", "rho_Y", "rho_D", "sigma_alpha_Z"),
  caption = "Pre-specified principal and stress grids."
)
```

# Phase map

```{r figure-phase-map, fig.cap="Phase map for whether the absolute Monte Carlo estimation bias is at least as large as the absolute mean specification shift caused by adding lagged Z. Arellano-Bond is excluded from this main comparison and retained as a sensitivity benchmark."}
phase_data <- summary_results |>
  dplyr::filter(
    design_family == "principal",
    estimator_id %in% c("fe_adl", "hpj_fe_adl")
  ) |>
  dplyr::mutate(
    phase = ifelse(bias_ge_abs_delta_z, "Bias >= |Delta_Z|", "Bias < |Delta_Z|"),
    rho_Y_label = paste0("rho_Y = ", rho_Y),
    N_label = paste0("N = ", N)
  )

phase_plot <- ggplot2::ggplot(
  phase_data,
  ggplot2::aes(x = factor(T), y = rho_Y_label, fill = phase)
) +
  ggplot2::geom_tile(color = "white", linewidth = 0.4) +
  ggplot2::facet_grid(estimator ~ N_label) +
  ggplot2::scale_fill_manual(values = c(
    "Bias >= |Delta_Z|" = "#C44E52",
    "Bias < |Delta_Z|" = "#4C78A8"
  )) +
  ggplot2::labs(
    x = "Number of observed periods (T)",
    y = "Outcome persistence",
    fill = "Finite-T comparison"
  ) +
  ggplot2::theme_minimal(base_size = 10) +
  ggplot2::theme(
    panel.grid = ggplot2::element_blank(),
    legend.position = "bottom"
  )

ggplot2::ggsave(
  filename = file.path(input_dir, "phase_map.pdf"),
  plot = phase_plot,
  width = 10,
  height = 5.5,
  units = "in"
)
ggplot2::ggsave(
  filename = file.path(input_dir, "phase_map.png"),
  plot = phase_plot,
  width = 10,
  height = 5.5,
  units = "in",
  dpi = 300
)
phase_plot
```

# Bias, RMSE, and coverage

```{r table-principal-results}
principal_table <- summary_results |>
  dplyr::filter(design_family == "principal") |>
  dplyr::group_by(estimator, N, T) |>
  dplyr::summarise(
    max_abs_bias = max(abs(bias), na.rm = TRUE),
    max_rmse = max(rmse, na.rm = TRUE),
    min_coverage = min(coverage, na.rm = TRUE),
    max_mcse_bias = max(mcse_bias, na.rm = TRUE),
    .groups = "drop"
  )
knitr::kable(
  principal_table,
  digits = 3,
  col.names = c("Estimator", "N", "T", "Max |bias|", "Max RMSE", "Min coverage", "Max bias MCSE"),
  caption = "Principal-grid bias, RMSE, coverage, and Monte Carlo precision summaries across outcome-persistence values."
)
```

# Arellano-Bond sensitivity diagnostics

Arellano-Bond is a sensitivity benchmark under its own moment conditions. It is not labeled as bias-corrected fixed effects. Instruments are collapsed and limited to lags 2--3; the DGP contains no common time shocks, so this benchmark uses individual effects in differences and does not add a growing set of time-dummy instruments.

```{r table-ab-diagnostics}
ab_table <- summary_results |>
  dplyr::filter(estimator_id == "arellano_bond") |>
  dplyr::group_by(design_family, N, T) |>
  dplyr::summarise(
    failure_rate = max(failure_rate, na.rm = TRUE),
    mean_instruments = max(mean_instrument_count, na.rm = TRUE),
    ar2_rejection_rate_05 = max(ar2_rejection_rate_05, na.rm = TRUE),
    hansen_rejection_rate_05 = max(hansen_rejection_rate_05, na.rm = TRUE),
    .groups = "drop"
  )
knitr::kable(
  ab_table,
  digits = 3,
  col.names = c("Design", "N", "T", "Failure", "Instruments", "AR(2) < .05", "Hansen < .05"),
  caption = "Arellano-Bond sensitivity-benchmark failure and diagnostic rejection rates."
)
```

# Monte Carlo stability and failures

```{r table-stability}
stability_table <- stability |>
  dplyr::group_by(estimator) |>
  dplyr::summarise(
    scenario_estimator_cells = dplyr::n(),
    cells_passing_all_stability_checks = sum(stability_pass, na.rm = TRUE),
    maximum_failure_rate = max(failure_rate, na.rm = TRUE),
    maximum_bias_mcse = max(mcse_bias, na.rm = TRUE),
    .groups = "drop"
  )
knitr::kable(
  stability_table,
  digits = 3,
  col.names = c("Estimator", "Cells", "Stable cells", "Max failure", "Max bias MCSE"),
  caption = "Monte Carlo precision, batch-stability, and estimator-failure checks."
)
```
