---
title: "Task 12 Static Audit: Simulation Boundary Conditions and Scope"
author: "Implementation audit"
date: "2026-07-11"
output:
  pdf_document:
    toc: true
    number_sections: true
geometry: margin=0.75in
fontsize: 10pt
header-includes:
  - \usepackage{pdflscape}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(data.table)
library(kableExtra)

audit_dir <- if (file.exists("task12_boundary_conditions.csv")) "." else "quality_reports"
boundary <- fread(file.path(audit_dir, "task12_boundary_conditions.csv"))
compact <- fread(file.path(audit_dir, "task12_boundary_conditions_compact.csv"))
reconciliation <- fread(file.path(audit_dir, "task12_reconciliation.csv"))
validation <- fread(file.path(audit_dir, "task12_static_validation.csv"))

make_longtable <- function(data, caption, widths, font_size = 7) {
  out <- knitr::kable(data, format = "latex", longtable = TRUE, booktabs = TRUE,
                      escape = TRUE, caption = caption)
  out <- kableExtra::column_spec(out, column = seq_along(widths), width = widths)
  kableExtra::kable_styling(out, latex_options = "repeat_header", font_size = font_size)
}
```

# Audit rule and scope

This is a static audit of already-versioned simulation scripts and result CSVs. It does not run or source any simulation, create any Monte Carlo draw, or modify a raw/result artifact. The companion script, `quality_reports/build_task12_boundary_conditions_audit.R`, reads the Task 10 inventory and existing CSVs and writes only the audit products under `quality_reports/`.

The evidentiary labels have a deliberately narrow meaning:

* **Existing result** means that Task 10 records both a source script and a declared versioned result artifact, and the Task 12 builder confirmed that both paths exist.
* **Declared output missing** means that the source script specifies an output path but the audited tree contains no corresponding versioned result. It cannot support a numerical conclusion.
* **Legacy artifact** means that a retained exploratory script has no persisted result artifact. It is not numerical evidence.

This audit reports the magnitudes stored in the existing result CSVs only to delimit the represented DGPs. A coefficient shift becomes causal bias only with the manuscript's DAG, timing, target estimand, and identification assumptions.

# Boundary-condition record

Table 1 records, family by family, what the existing code actually varies and holds fixed; what the retained artifacts measure; what remains unidentified; provenance; replication counts; and the appropriate interpretation boundary. A requested replication count is not silently substituted for the count actually persisted in the summary CSV when unstable draws were discarded.

```{=latex}
\begin{landscape}
```

```{r boundary-table, results='asis'}
boundary_detail <- boundary[, .(
  `Family` = family,
  `Design: varied versus fixed` = paste0("Varied: ", actually_varied, " Fixed: ", held_fixed),
  `Evidence, magnitude, and replications` = paste0("Status: ", evidence_status, ". ", measured, " Replications: ", replication_count),
  `Not identified and interpretation boundary` = paste0(not_identified, " Scope: ", interpretation_boundary),
  `Task 10 inventory IDs` = inventory_ids
)]

make_longtable(
  boundary_detail,
  "Static boundary-condition record for the existing simulation artifacts.",
  c("2.7cm", "4.8cm", "5.0cm", "5.7cm", "4.4cm"),
  font_size = 5
)
```

```{=latex}
\end{landscape}
```

# Reader-facing compact table

The manuscript uses the following compact version after Figure 2. It is a map of what the reported simulations delimit, not a claim that the conditional baseline is generally robust.

```{r compact-table, results='asis'}
compact_out <- knitr::kable(
  compact,
  format = "latex",
  booktabs = TRUE,
  longtable = TRUE,
  escape = TRUE,
  caption = paste0(
    "Boundary conditions represented by existing Monte Carlo artifacts. Each row is a design-specific result, ",
    "not a general robustness or identification claim. Exact grids and provenance are in the Task 12 audit."
  )
)
compact_out <- kableExtra::column_spec(compact_out, 1, width = "3.1cm")
compact_out <- kableExtra::column_spec(compact_out, 2, width = "6.1cm")
compact_out <- kableExtra::column_spec(compact_out, 3, width = "5.4cm")
kableExtra::kable_styling(compact_out, latex_options = "repeat_header", font_size = 7)
```

# Reconciliation with the Task 10 inventory and manuscript

```{r reconciliation-table, results='asis'}
make_longtable(
  reconciliation,
  "Reconciliation of the Task 10 inventory with the current PA manuscript claims.",
  c("4.4cm", "8.5cm", "3.2cm"),
  font_size = 7
)
```

The material correction is negative rather than additive: no numerical finite-\( T \) conclusion can be drawn from `sim_dual_role_z_varyT.R`, because its declared 200-rep output is absent. The existing eight-model finite-\( T \) result is distinct, has a saved CSV, and keeps \( N=100 \) and burn-in 100 fixed. Similarly, `sim_ivb_completa.R` remains a legacy exploratory source without a persisted result and is excluded from numerical support.

# Validation record

```{r validation-table, results='asis'}
knitr::kable(validation, format = "latex", booktabs = TRUE,
             caption = "Static validation checks performed by the Task 12 builder.") %>%
  kableExtra::kable_styling(latex_options = "hold_position")
```

The complete source-script paths and result-artifact paths are retained, without abbreviation, in `quality_reports/task12_boundary_conditions.csv`; the concise Task 10 inventory identifiers in Table 1 keep the rendered audit legible. The protected simulation directory is checked again after rendering against the SHA-256 manifest created before this audit. Rendering this report reads only the generated Task 12 tables and does not execute a simulation.
