---
title: "Included Variable Bias in Ballard-Rosa, Mosley & Wellhausen (2022, IO): An Empirical Application"
author: ""
date: "`r Sys.Date()`"
output:
  pdf_document:
    number_sections: true
    keep_tex: false
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 5
)
library(fixest)
library(dplyr)
library(knitr)
source("../../ivb_utils.R")
```

# Introduction

Ballard-Rosa, Mosley, and Wellhausen (2022, *International Organization*) estimate the effect of government partisanship on the proportion of sovereign bonds issued in domestic currency. The main specification (Table 2, Column 3) is a country fixed effects model estimated on non-OECD countries at monthly frequency:

$$
\text{propDom}_{it} = \beta_R \, \text{Right}_{i,t-1} + \beta_L \, \text{Left}_{i,t-1} + \mathbf{Z}'_{it}\gamma + f(t) + \alpha_i + \varepsilon_{it}
$$

where $\text{propDom}_{it}$ is the proportion of sovereign bonds with maturity greater than 1 year issued in domestic currency (0--1), $\text{Right}_{i,t-1}$ and $\text{Left}_{i,t-1}$ are lagged partisan dummies (with centrist/other as the base category), $\mathbf{Z}_{it}$ is a vector of 16 controls (most lagged 12 months), $f(t)$ is a cubic time polynomial, and $\alpha_i$ are country fixed effects. Standard errors are clustered by country.

The paper's key finding is that left-wing governments issue significantly more domestic-currency debt ($\hat\beta_L = 0.109$, $p < 0.05$), while right-wing governments issue less ($\hat\beta_R = -0.075$, $p < 0.05$).

A key question is whether any of these controls are **colliders** --- variables caused by both the treatment (partisanship) and the outcome (bond denomination choice). If so, conditioning on them introduces **Included Variable Bias** (IVB). The IVB for each candidate collider $z_j$ is given by:

$$
\text{IVB}_j(\hat\beta_L) = \hat\beta_L^{\text{short}_j} - \hat\beta_L^{\text{long}} = -\hat\theta_j \cdot \hat\pi_j
$$

where $\hat\theta_j$ is the coefficient of $z_j$ in the "long" model (with all controls), and $\hat\pi_j$ is the coefficient of $\text{Left}_{i,t-1}$ in the auxiliary regression of $z_j$ on the treatment dummies and remaining controls.

Since the model includes **two treatment variables** (Right and Left, with Center as base), the IVB decomposition applies separately to each treatment coefficient. We focus on $\hat\beta_L$ (the Left coefficient) as it represents the paper's main finding.

# Data

```{r load-data}
dat <- read.delim(
  "Coming_to_Terms_data.tab",
  stringsAsFactors = FALSE
)
cat("Raw data:", nrow(dat), "rows,", ncol(dat), "columns\n")
cat("Countries:", length(unique(dat$ccode)), "\n")
cat("Non-OECD countries:", length(unique(dat$ccode[dat$oecd == 0])), "\n")
```

## Variable preparation

The Stata specification uses `l12.` (12-month lag) for most controls and `l.` (1-month lag) for the treatment. We create these lags in R by sorting within country-time panels. The data variables `oil_rents_gdp` and `fdi_net_inflow_gdp` correspond to the Stata do-file's `oil_rents` and `fdi_net_inflow` respectively.

```{r create-lags}
dat <- dat %>%
  arrange(ccode, time) %>%
  group_by(ccode) %>%
  mutate(
    # Treatment: 1-month lag of partisan dummies
    l1_rightwing = lag(rightwing_exec_mo, 1),
    l1_leftwing  = lag(leftwing_exec_mo, 1),
    # Controls: 12-month lags
    l12_lngdppc       = lag(lngdppc, 12),
    l12_gdp_growth    = lag(gdp_growth, 12),
    l12_avgDebt_gdp   = lag(avgDebt_gdp, 12),
    l12_curr_act_gdp  = lag(curr_act_gdp, 12),
    l12_tradeGDP      = lag(tradeGDP, 12),
    l12_oil_rents     = lag(oil_rents_gdp, 12),
    l12_fdi_net       = lag(fdi_net_inflow_gdp, 12),
    l12_treasury10yr  = lag(treasury10yr, 12),
    l12_peg           = lag(peg, 12),
    l12_highCBI       = lag(highCBI, 12),
    l12_kaopen        = lag(kaopen, 12),
    l12_imfAnyInPlace = lag(imfAnyInPlace, 12),
    l12_v2x_polyarchy = lag(v2x_polyarchy, 12)
  ) %>%
  ungroup()

# Crisis variables are NOT lagged (contemporaneous)
# They already exist as: crisis_currency, crisis_inflation, crisis_sovdebt

# Filter to non-OECD
dat_nooecd <- dat %>% filter(oecd == 0)
cat("Non-OECD observations:", nrow(dat_nooecd), "\n")
```

# Specification

The Stata command for Table 2, Column 3 is:

```
xtreg propDom_gt1yr l.ib2.execrlc_mo l12.lngdppc l12.gdp_growth
  l12.avgDebt_gdp l12.curr_act_gdp l12.tradeGDP l12.oil_rents
  l12.fdi_net_inflow l12.treasury10yr l12.peg l12.highCBI l12.kaopen
  l12.imfAnyInPlace crisis_currency crisis_inflation crisis_sovdebt
  l12.v2x_polyarchy time time2 time3
  if oecd == 0, fe vce(cluster ccode)
```

The variables are:

| Role | Variable | Description |
|------|----------|-------------|
| $Y$ | `propDom_gt1yr` | Proportion of bonds > 1yr maturity issued in domestic currency |
| $D_R$ | `l.rightwing_exec_mo` | Right government (1-month lag) |
| $D_L$ | `l.leftwing_exec_mo` | Left government (1-month lag) |
| $Z_1$ | `l12.lngdppc` | Log GDP per capita (12-month lag) |
| $Z_2$ | `l12.gdp_growth` | GDP growth (12-month lag) |
| $Z_3$ | `l12.avgDebt_gdp` | External debt / GDP (12-month lag) |
| $Z_4$ | `l12.curr_act_gdp` | Current account / GDP (12-month lag) |
| $Z_5$ | `l12.tradeGDP` | Trade / GDP (12-month lag) |
| $Z_6$ | `l12.oil_rents` | Oil rents / GDP (12-month lag) |
| $Z_7$ | `l12.fdi_net_inflow` | FDI net inflows / GDP (12-month lag) |
| $Z_8$ | `l12.treasury10yr` | US 10-year Treasury rate (12-month lag) |
| $Z_9$ | `l12.peg` | Pegged exchange rate (12-month lag) |
| $Z_{10}$ | `l12.highCBI` | High central bank independence (12-month lag) |
| $Z_{11}$ | `l12.kaopen` | Chinn-Ito capital account openness (12-month lag) |
| $Z_{12}$ | `l12.imfAnyInPlace` | IMF program in place (12-month lag) |
| $Z_{13}$ | `crisis_currency` | Currency crisis (contemporaneous) |
| $Z_{14}$ | `crisis_inflation` | Inflation crisis (contemporaneous) |
| $Z_{15}$ | `crisis_sovdebt` | Sovereign debt crisis (contemporaneous) |
| $Z_{16}$ | `l12.v2x_polyarchy` | V-Dem democracy index (12-month lag) |
| Temporal | `time`, `time2`, `time3` | Cubic polynomial in time |
| FE | `ccode` | Country fixed effects |

# Replication of Table 2, Column 3

```{r replication}
mod_full <- feols(
  propDom_gt1yr ~ l1_rightwing + l1_leftwing +
    l12_lngdppc + l12_gdp_growth + l12_avgDebt_gdp + l12_curr_act_gdp +
    l12_tradeGDP + l12_oil_rents + l12_fdi_net + l12_treasury10yr +
    l12_peg + l12_highCBI + l12_kaopen + l12_imfAnyInPlace +
    crisis_currency + crisis_inflation + crisis_sovdebt +
    l12_v2x_polyarchy + time + time2 + time3 | ccode,
  data = dat_nooecd, vcov = ~ccode
)
```

```{r replication-table}
published <- c(
  l1_rightwing     = -0.075,
  l1_leftwing      =  0.109,
  l12_lngdppc      = -0.047,
  l12_gdp_growth   =  0.002,
  l12_avgDebt_gdp  =  0.001,
  l12_curr_act_gdp =  0.003,
  l12_tradeGDP     = -0.000,
  l12_oil_rents    = -0.002,
  l12_fdi_net      =  0.001,
  l12_treasury10yr =  0.022,
  l12_peg          = -0.006,
  l12_highCBI      =  0.011,
  l12_kaopen       =  0.005,
  l12_imfAnyInPlace = -0.045,
  crisis_currency  = -0.010,
  crisis_inflation = -0.223,
  crisis_sovdebt   =  0.102,
  l12_v2x_polyarchy = -0.012
)

published_se <- c(
  l1_rightwing     = 0.033,
  l1_leftwing      = 0.044,
  l12_lngdppc      = 0.052,
  l12_gdp_growth   = 0.002,
  l12_avgDebt_gdp  = 0.001,
  l12_curr_act_gdp = 0.003,
  l12_tradeGDP     = 0.001,
  l12_oil_rents    = 0.006,
  l12_fdi_net      = 0.002,
  l12_treasury10yr = 0.008,
  l12_peg          = 0.032,
  l12_highCBI      = 0.031,
  l12_kaopen       = 0.014,
  l12_imfAnyInPlace = 0.028,
  crisis_currency  = 0.021,
  crisis_inflation = 0.056,
  crisis_sovdebt   = 0.074,
  l12_v2x_polyarchy = 0.137
)

var_labels <- c(
  "Right government", "Left government",
  "Log GDP p.c.", "GDP growth", "External debt/GDP", "Current account/GDP",
  "Trade/GDP", "Oil rents/GDP", "FDI/GDP", "US Treasury 10yr",
  "Pegged XR", "High CBI", "Chinn-Ito openness", "IMF program",
  "Currency crisis", "Inflation crisis", "Sov. debt crisis",
  "Democracy (V-Dem)"
)

rep_coefs <- coef(mod_full)[names(published)]
rep_se <- sqrt(diag(vcov(mod_full)))[names(published)]

comparison <- data.frame(
  Variable     = var_labels,
  Pub_Coef     = published,
  Rep_Coef     = round(as.numeric(rep_coefs), 3),
  Pub_SE       = published_se,
  Rep_SE       = round(as.numeric(rep_se), 3),
  stringsAsFactors = FALSE
)

kable(comparison, digits = 3, row.names = FALSE,
      caption = "Replication of Table 2, Column 3 (Full Controls, non-OECD)",
      col.names = c("Variable", "Published Coef", "Replicated Coef",
                     "Published SE", "Replicated SE"))
```

```{r model-details}
n_countries <- length(fixest::fixef(mod_full)$ccode)
cat("N =", mod_full$nobs, "\n")
cat("Countries =", n_countries, "\n")
cat("R-squared (within) =", round(r2(mod_full, "wr2"), 3), "\n")
```

The replicated Left government coefficient is $\hat\beta_L$ = `r round(coef(mod_full)["l1_leftwing"], 3)` (SE = `r round(sqrt(diag(vcov(mod_full)))["l1_leftwing"], 3)`), compared to the published value of 0.109 (0.044). Minor numerical differences may arise from Stata's `xtreg, fe` degrees-of-freedom adjustments versus `fixest`'s defaults.

# IVB Decomposition

## Setup

We define:

- **Treatment** ($D$): `l1_rightwing` and `l1_leftwing` (lagged partisan dummies, Center is base)
- **Outcome** ($Y$): `propDom_gt1yr` (proportion of bonds > 1yr in domestic currency)
- **Candidate colliders** ($z_j$): each of the 16 controls, one at a time
- **Temporal covariates** (always in $w$): `time`, `time2`, `time3`
- **Fixed effects**: `ccode` (country)

For each candidate collider $z_j$, the "short" model excludes $z_j$ (but retains the other 15 controls plus temporal covariates), and the "long" model includes all 16 controls plus temporal covariates. The IVB identity $\hat\beta_L^{\text{long}} - \hat\beta_L^{\text{short}_j} = -\hat\theta_j \cdot \hat\pi_j$ is verified numerically.

Since `compute_ivb_multi()` supports multiple treatment variables in `d_vars`, we pass both `l1_rightwing` and `l1_leftwing` and extract the results for the Left coefficient.

## Decomposition

```{r ivb-decomposition}
d_vars <- c("l1_rightwing", "l1_leftwing")
temporal_vars <- c("time", "time2", "time3")

all_controls <- c(
  "l12_lngdppc", "l12_gdp_growth", "l12_avgDebt_gdp", "l12_curr_act_gdp",
  "l12_tradeGDP", "l12_oil_rents", "l12_fdi_net", "l12_treasury10yr",
  "l12_peg", "l12_highCBI", "l12_kaopen", "l12_imfAnyInPlace",
  "crisis_currency", "crisis_inflation", "crisis_sovdebt",
  "l12_v2x_polyarchy"
)

control_labels <- c(
  "Log GDP p.c.", "GDP growth", "External debt/GDP", "Current account/GDP",
  "Trade/GDP", "Oil rents/GDP", "FDI/GDP", "US Treasury 10yr",
  "Pegged XR", "High CBI", "Chinn-Ito openness", "IMF program",
  "Currency crisis", "Inflation crisis", "Sov. debt crisis",
  "Democracy (V-Dem)"
)

ivb_results <- lapply(seq_along(all_controls), function(j) {
  z_var <- all_controls[j]
  w_vars <- c(setdiff(all_controls, z_var), temporal_vars)
  compute_ivb_multi(
    data = dat_nooecd, y = "propDom_gt1yr", d_vars = d_vars,
    z = z_var, w = w_vars,
    fe = "ccode", vcov = "iid"
  )
})
names(ivb_results) <- all_controls

# Extract results for Left coefficient (l1_leftwing)
tab_ivb <- do.call(rbind, lapply(seq_along(all_controls), function(j) {
  r <- ivb_results[[j]]$results
  r_left <- r[r$term == "l1_leftwing", ]
  r_right <- r[r$term == "l1_rightwing", ]
  data.frame(
    Collider      = control_labels[j],
    Variable      = all_controls[j],
    beta_short_L  = r_left$beta_short,
    beta_long_L   = r_left$beta_long,
    theta         = r_left$theta,
    pi_L          = r_left$pi,
    IVB_L         = r_left$ivb_formula,
    check_L       = r_left$diff_check,
    beta_short_R  = r_right$beta_short,
    beta_long_R   = r_right$beta_long,
    pi_R          = r_right$pi,
    IVB_R         = r_right$ivb_formula,
    check_R       = r_right$diff_check,
    stringsAsFactors = FALSE
  )
}))
```

## Results for Left Government Coefficient

```{r ivb-table-left}
tab_left <- tab_ivb[, c("Collider", "beta_short_L", "beta_long_L",
                          "theta", "pi_L", "IVB_L", "check_L")]
tab_left$pct_beta <- round(100 * tab_left$IVB_L / tab_left$beta_long_L, 2)

kable(tab_left, digits = 6, row.names = FALSE,
      caption = "IVB decomposition for Left government coefficient (Table 2, Col 3)",
      col.names = c("Candidate collider $z_j$",
                     "$\\hat\\beta_L^{\\text{short}_j}$",
                     "$\\hat\\beta_L^{\\text{long}}$",
                     "$\\hat\\theta_j$",
                     "$\\hat\\pi_j^L$",
                     "IVB $= -\\hat\\theta_j \\hat\\pi_j^L$",
                     "Check",
                     "IVB / $\\hat\\beta_L^{\\text{long}}$ (\\%)"))
```

The identity holds exactly (check $\approx 0$) in all cases, confirming the numerical validity of the decomposition.

## Results for Right Government Coefficient

```{r ivb-table-right}
tab_right <- tab_ivb[, c("Collider", "beta_short_R", "beta_long_R",
                           "theta", "pi_R", "IVB_R", "check_R")]
tab_right$pct_beta <- round(100 * tab_right$IVB_R / tab_right$beta_long_R, 2)

kable(tab_right, digits = 6, row.names = FALSE,
      caption = "IVB decomposition for Right government coefficient (Table 2, Col 3)",
      col.names = c("Candidate collider $z_j$",
                     "$\\hat\\beta_R^{\\text{short}_j}$",
                     "$\\hat\\beta_R^{\\text{long}}$",
                     "$\\hat\\theta_j$",
                     "$\\hat\\pi_j^R$",
                     "IVB $= -\\hat\\theta_j \\hat\\pi_j^R$",
                     "Check",
                     "IVB / $\\hat\\beta_R^{\\text{long}}$ (\\%)"))
```

## Summary of IVB Components (Left Coefficient)

```{r ivb-summary}
tab_summary <- data.frame(
  Control   = tab_ivb$Collider,
  theta     = round(tab_ivb$theta, 4),
  pi_L      = round(tab_ivb$pi_L, 4),
  IVB_L     = round(tab_ivb$IVB_L, 6),
  pct       = round(100 * tab_ivb$IVB_L / tab_ivb$beta_long_L, 2),
  Direction = ifelse(tab_ivb$IVB_L > 0, "Inflates", "Deflates")
)

kable(tab_summary, digits = 4, row.names = FALSE,
      caption = "Summary of IVB components for Left government coefficient",
      col.names = c("Control", "$\\hat\\theta_j$ (effect on $Y$)",
                     "$\\hat\\pi_j^L$ (effect of Left on $Z_j$)",
                     "IVB", "IVB/$\\hat{\\beta}_L^{\\text{long}}$ (\\%)",
                     "Direction"))
```

## Visualization

```{r ivb-barplot, fig.height=6}
# Sort by absolute IVB
ord <- order(abs(tab_ivb$IVB_L), decreasing = TRUE)
ivb_sorted <- tab_ivb$IVB_L[ord]
labels_sorted <- tab_ivb$Collider[ord]

par(mar = c(5, 10, 4, 2))
cols <- ifelse(ivb_sorted > 0, "steelblue", "firebrick")
barplot(ivb_sorted, names.arg = labels_sorted, horiz = TRUE, las = 1,
        col = cols, border = NA,
        main = "IVB Decomposition for Left Government Coefficient",
        xlab = expression(IVB[j] == -hat(theta)[j] %.% hat(pi)[j]^L),
        cex.names = 0.75)
abline(v = 0, lty = 2, col = "gray40")
legend("topright",
       legend = c("Inflates (IVB > 0)", "Deflates (IVB < 0)"),
       fill = c("steelblue", "firebrick"), border = NA, cex = 0.8)
```

## Total IVB and No-Controls Benchmark

```{r total-ivb}
total_ivb_L <- sum(tab_ivb$IVB_L)
beta_long_L <- tab_ivb$beta_long_L[1]

# Model with NO controls (only temporal + FE)
# Use the same sample as the full model for comparability
all_vars <- c("propDom_gt1yr", d_vars, all_controls, temporal_vars, "ccode")
dat_complete <- dat_nooecd[complete.cases(dat_nooecd[, all_vars]), ]

mod_no_controls <- feols(
  propDom_gt1yr ~ l1_rightwing + l1_leftwing +
    time + time2 + time3 | ccode,
  data = dat_complete, vcov = "iid"
)
beta_no_controls_L <- coef(mod_no_controls)["l1_leftwing"]
```

- $\hat\beta_L^{\text{long}}$ (with all controls) = `r round(beta_long_L, 4)`
- $\hat\beta_L^{\text{no controls}}$ (only temporal + FE, same sample) = `r round(beta_no_controls_L, 4)`
- **Total IVB** (sum across all controls) = `r round(total_ivb_L, 6)` (`r round(100 * total_ivb_L / beta_long_L, 1)`% of $\hat\beta_L^{\text{long}}$)

**Note:** The total IVB (sum of individual IVBs) does not generally equal $\hat\beta_L^{\text{long}} - \hat\beta_L^{\text{no controls}}$ because the IVB decomposition drops one control at a time, not all simultaneously. The sum provides a useful summary of the total sensitivity to control specification.

## Top Contributors

```{r top-contributors}
tab_ranked <- tab_ivb[order(abs(tab_ivb$IVB_L), decreasing = TRUE), ]
tab_top <- data.frame(
  Rank    = 1:nrow(tab_ranked),
  Control = tab_ranked$Collider,
  IVB     = round(tab_ranked$IVB_L, 6),
  Pct     = round(100 * tab_ranked$IVB_L / tab_ranked$beta_long_L, 2),
  theta   = round(tab_ranked$theta, 4),
  pi_L    = round(tab_ranked$pi_L, 4),
  stringsAsFactors = FALSE
)
kable(tab_top, digits = 4, row.names = FALSE,
      caption = "Controls ranked by absolute IVB for Left coefficient",
      col.names = c("Rank", "Control", "IVB",
                     "IVB/$\\hat{\\beta}_L$ (\\%)",
                     "$\\hat\\theta_j$", "$\\hat\\pi_j^L$"))
```

# Literature Support for Collider Status

The IVB formula diagnoses **how much** each control shifts the treatment estimate, but **whether** a control is a collider depends on substantive knowledge: does government partisanship cause the control (treatment $\to$ $Z_j$), and does bond denomination choice also cause or respond to the control ($Y \to Z_j$)?

We review the literature for the controls with the largest IVB magnitudes.

## Partisanship $\to$ Inflation Crisis ($Z_{14}$)

There is strong evidence linking government ideology to inflation outcomes:

- **Hibbs (1977, APSR):** Left governments tolerate higher inflation to achieve lower unemployment, the classic partisan theory of macroeconomic policy.
- **Alesina (1987, QJE):** Rational partisan theory predicts temporary inflation increases after left-wing electoral victories.
- **Broz (2002, IO):** Left governments in developing countries have less anti-inflation credibility, potentially triggering inflation crises.

At the same time, inflation crises directly affect the currency denomination of sovereign bonds (investors demand hard-currency denomination during inflationary episodes), establishing a plausible $Y \to Z$ link.

**Implication:** Inflation crisis is a plausible collider. The very large $\hat\theta_{14}$ (inflation crisis strongly reduces domestic-currency issuance) combined with partisan differences in inflation risk creates IVB.

## Partisanship $\to$ IMF Program ($Z_{12}$)

There is moderate evidence that partisanship affects IMF program participation:

- **Vreeland (2003, Cambridge UP):** Governments may strategically enter IMF programs based on political ideology and domestic political constraints.
- **Dreher (2006, JDE):** Left-wing governments may be more reluctant to accept IMF conditionality, while fiscal crises under either ideology can trigger IMF involvement.

IMF programs may also influence bond denomination through conditionality requiring macroeconomic reforms.

## Partisanship $\to$ Capital Account Openness ($Z_{11}$)

- **Quinn & Toyoda (2008, AJPS):** Partisan governments differ in their preferences for capital account liberalization, with right-wing governments generally more favorable.
- **Bearce (2003, IO):** The political economy of capital mobility reflects partisan preferences over monetary policy autonomy.

## Partisanship $\to$ GDP per capita ($Z_1$) and Growth ($Z_2$)

- **Alesina, Roubini & Cohen (1997):** Partisan effects on macroeconomic outcomes are well-documented but modest and temporary.
- The 12-month lag substantially mitigates contemporaneous collider concerns for these variables, as partisan effects on GDP levels take time to materialize.

## Partisanship $\to$ Exchange Rate Regime ($Z_9$)

- **Broz & Frieden (2001, IO):** Exchange rate regime choice reflects political economy considerations, with left governments potentially preferring flexible rates for monetary policy autonomy.
- **Bearce & Hallerberg (2011, IO):** Partisan preferences shape exchange rate policy in open economies.

## Partisanship $\to$ Current Account ($Z_4$) and Trade ($Z_5$)

- **Milner & Judkins (2004, IO):** Left parties in developing countries may adopt different trade policies, affecting trade openness and current account balances.
- The 12-month lag again attenuates contemporaneous collider concerns.

# Summary

```{r summary-table-final}
# Classify controls by IVB magnitude
top5 <- tab_ivb[order(abs(tab_ivb$IVB_L), decreasing = TRUE), ][1:5, ]

summary_all <- data.frame(
  Control = tab_ivb$Collider,
  `D causes Z?` = c(
    "Weak (lagged 12mo)", "Weak (lagged 12mo)",      # GDP p.c., GDP growth
    "Weak", "Weak",                                    # External debt, Current account
    "Possible", "Weak",                                # Trade, Oil rents
    "Weak", "No (exogenous)",                          # FDI, US Treasury
    "Possible", "Weak",                                # Peg, CBI
    "Possible", "Possible",                            # Chinn-Ito, IMF
    "Weak", "Strong (partisan theory)",                # Currency crisis, Inflation crisis
    "Weak", "Possible"                                 # Sov debt crisis, Democracy
  ),
  `Y causes Z?` = c(
    "Weak", "Weak",                                    # GDP p.c., GDP growth
    "Possible", "Possible",                            # External debt, Current account
    "Weak", "No",                                      # Trade, Oil rents
    "Possible", "No (exogenous)",                      # FDI, US Treasury
    "Possible", "Weak",                                # Peg, CBI
    "Possible", "Possible",                            # Chinn-Ito, IMF
    "Possible", "Strong",                              # Currency crisis, Inflation crisis
    "Possible", "Weak"                                 # Sov debt crisis, Democracy
  ),
  `Collider?` = c(
    "Unlikely", "Unlikely",
    "Unlikely", "Unlikely",
    "Unlikely", "Unlikely",
    "Unlikely", "No",
    "Possible", "Unlikely",
    "Possible", "Possible",
    "Possible", "Plausible",
    "Possible", "Unlikely"
  ),
  IVB = round(tab_ivb$IVB_L, 6),
  `IVB pct` = round(100 * tab_ivb$IVB_L / tab_ivb$beta_long_L, 2),
  check.names = FALSE, stringsAsFactors = FALSE
)

kable(summary_all, digits = 4, row.names = FALSE,
      caption = "Summary: Literature evidence and IVB magnitudes for Left coefficient",
      col.names = c("Control", "$D \\to Z$?", "$Y \\to Z$?",
                     "Collider?", "IVB",
                     "IVB/$\\hat{\\beta}_L$ (\\%)"))
```

## Key Findings

```{r key-findings}
# Identify the top 3 contributors
top3 <- tab_ivb[order(abs(tab_ivb$IVB_L), decreasing = TRUE), ][1:3, ]
cat("Top 3 IVB contributors (by absolute magnitude):\n")
for (i in 1:3) {
  cat(sprintf("  %d. %s: IVB = %.6f (%.1f%% of beta_L)\n",
              i, top3$Collider[i], top3$IVB_L[i],
              100 * top3$IVB_L[i] / top3$beta_long_L[i]))
}
```

The IVB decomposition for the Left government coefficient ($\hat\beta_L$ = `r round(beta_long_L, 3)`) reveals the following:

1. **Most controls produce small IVBs.** This is consistent with the fact that government partisanship (lagged 1 month) has limited correlation with most economic controls (lagged 12 months) conditional on country fixed effects. The temporal separation between treatment and controls reduces the scope for collider bias.

2. **Inflation crisis** is the most substantively concerning potential collider. Partisan theory (Hibbs 1977; Alesina 1987) predicts that left-wing governments face higher inflation risk, and inflation crises directly reduce domestic-currency bond issuance (the large negative $\hat\theta$). If left-wing governance causes inflation crises (positive $\hat\pi_L$), conditioning on inflation crisis absorbs part of the negative pathway from Left $\to$ inflation crisis $\to$ less domestic-currency debt, potentially biasing $\hat\beta_L$ upward.

3. **The US 10-year Treasury rate** is exogenous to individual-country partisanship and bond denomination, so it cannot be a collider. Any IVB from this variable reflects mechanical omitted-variable-bias correction, not collider bias.

4. **The 12-month lag structure** of most controls substantially mitigates collider concerns. For partisanship to "cause" a 12-month lagged control conditional on country fixed effects, it would need to affect the control with a lead of at least 12 months, which is implausible for most macroeconomic variables in a monthly panel.

5. **The total IVB** is `r round(total_ivb_L, 4)` (`r round(100 * total_ivb_L / beta_long_L, 1)`% of $\hat\beta_L^{\text{long}}$), suggesting that the overall sensitivity of the Left coefficient to the full control specification is modest.

# References

- Alesina, A. (1987). Macroeconomic Policy in a Two-Party System as a Repeated Game. *Quarterly Journal of Economics*, 102(3), 651--678.
- Alesina, A., Roubini, N., & Cohen, G. D. (1997). *Political Cycles and the Macroeconomy*. MIT Press.
- Ballard-Rosa, C., Mosley, L., & Wellhausen, R. L. (2022). Coming to Terms: The Politics of Sovereign Bond Denomination. *International Organization*, 76(1), 126--153.
- Bearce, D. H. (2003). Societal Preferences, Partisan Agents, and Monetary Policy Outcomes. *International Organization*, 57(2), 373--410.
- Bearce, D. H. & Hallerberg, M. (2011). Democracy and De Facto Exchange Rate Regimes. *Economics & Politics*, 23(2), 172--194.
- Broz, J. L. (2002). Political System Transparency and Monetary Commitment Regimes. *International Organization*, 56(4), 861--887.
- Broz, J. L. & Frieden, J. A. (2001). The Political Economy of International Monetary Relations. *Annual Review of Political Science*, 4, 317--343.
- Dreher, A. (2006). IMF and Economic Growth: The Effects of Programs, Loans, and Compliance with Conditionality. *World Development*, 34(5), 769--788.
- Hibbs, D. A. (1977). Political Parties and Macroeconomic Policy. *American Political Science Review*, 71(4), 1467--1487.
- Milner, H. V. & Judkins, B. (2004). Partisanship, Trade Policy, and Globalization: Is There a Left-Right Divide on Trade Policy? *International Studies Quarterly*, 48(1), 95--119.
- Quinn, D. P. & Toyoda, A. M. (2008). Does Capital Account Liberalization Lead to Growth? *Review of Financial Studies*, 21(3), 1403--1449.
- Vreeland, J. R. (2003). *The IMF and Economic Development*. Cambridge University Press.
