---
title: "Allocating Specification Shifts with Correlated Controls"
subtitle: "One-at-a-time, conditional leave-one-out, and Shapley diagnostics"
author: "Online appendix for the IVB paper"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    latex_engine: xelatex
    number_sections: true
    toc: true
    keep_tex: false
header-includes:
  - \usepackage{booktabs}
  - \usepackage{float}
  - \usepackage{geometry}
  - \geometry{margin=1in}
fontsize: 11pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
source("scripts/shift_allocation_correlated_controls.R")
```

# Purpose and scope

When candidate controls are correlated, the joint specification shift is unique but its control-level allocation is not. Let $\beta(S)$ denote the coefficient on the focal treatment in the linear model containing the common regressors $W$ and the subset $S$ of candidate controls, all estimated on the same observations. Define

\[
\Delta(S)=\beta(S)-\beta(\varnothing),
\qquad
\Delta_{\mathcal Z}=\beta(\mathcal Z)-\beta(\varnothing).
\]

This appendix compares three descriptive diagnostics. Each is a contrast between nested linear projections; none establishes causal bias relative to the contemporaneous treatment effect (CET). In a linear TSCS application, $W$ can include unit and time fixed effects, lagged outcomes, lagged treatments, and other defended lagged state variables.

# Three diagnostics

For $j\in\mathcal Z$, the **one-at-a-time** increment is

\[
a_j=\beta(\{j\})-\beta(\varnothing).
\]

It answers how much the coefficient moves when $Z_j$ alone is added to the short model. It holds every other candidate control out, so with correlated controls the values $a_j$ need not add to the joint shift.

The **leave-one-control-out** (LOO) conditional increment is

\[
\ell_j=\beta(\mathcal Z)-\beta(\mathcal Z\setminus\{j\}).
\]

It is the increment from adding $Z_j$ last to a model that already contains all other candidates. Equivalently, deleting $Z_j$ from the full model changes the coefficient by $-\ell_j$. Thus $\ell_j$ is a last-in conditional sensitivity, not a share of $\Delta_{\mathcal Z}$; across $j$, these increments need not add to the joint shift.

For an inclusion order $\sigma$, write $P_j^\sigma$ for the controls preceding $j$. The path-specific sequential increment is

\[
c_j^\sigma=\beta(P_j^\sigma\cup\{j\})-\beta(P_j^\sigma).
\]

It telescopes along each path: $\sum_j c_j^\sigma=\Delta_{\mathcal Z}$. The **Shapley value** averages this increment over all $q!$ inclusion orders,

\[
\phi_j=\frac{1}{q!}\sum_{\sigma}c_j^\sigma.
\]

Consequently, $\sum_j\phi_j=\Delta_{\mathcal Z}$. This symmetric allocation averages shared variation across possible orders. It is an accounting convention rather than a causal decomposition, and its factorial enumeration becomes costly as the number of candidate controls grows.

# Deterministic correlated-control example

The following example is identical in construction to the deterministic fixture used for the multivariate joint-shift derivation. It uses no random-number generation and no paper simulation. All models use the same 240 observations and unweighted OLS.

```{r deterministic-design, include=FALSE}
n <- 240L
index <- seq_len(n)
w1 <- as.numeric(scale(index))
w2 <- as.numeric(scale((index - mean(index))^2))
d <- 0.55 * w1 - 0.20 * w2 + sin(index / 7) + 0.35 * cos(index / 13)

u1 <- sin(index / 3) + 0.20 * cos(index / 17)
u2 <- cos(index / 5) - 0.15 * sin(index / 19)
u3 <- sin(index / 11) + 0.25 * cos(index / 23)
z1 <- 0.90 * d + 0.55 * w1 + 0.45 * u1
z2 <- -0.45 * d + 0.35 * w2 + 0.70 * z1 + 0.35 * u2
z3 <- 0.55 * d - 0.30 * w1 + 0.40 * z1 - 0.35 * z2 + 0.30 * u3
y <- 1.25 * d + 0.80 * z1 - 0.55 * z2 + 0.65 * z3 +
  0.40 * w1 - 0.25 * w2 + 0.20 * sin(index / 2) - 0.10 * cos(index / 29)

W <- cbind(W1 = w1, W2 = w2)
Z <- cbind(Z1 = z1, Z2 = z2, Z3 = z3)
joint_shift <- shift_set(y, d, W, Z, colnames(Z))
one_results <- one_at_a_time(y, d, W, Z)
loo_results <- leave_one_out(y, d, W, Z)
shapley_results <- shapley_allocation(y, d, W, Z)
diagnostic_table <- Reduce(
  function(left, right) merge(left, right, by = "control", sort = FALSE),
  list(one_results, loo_results, shapley_results)
)
diagnostic_table <- diagnostic_table[match(colnames(Z), diagnostic_table$control), ]
```

```{r diagnostics-table}
diagnostic_display <- diagnostic_table
names(diagnostic_display) <- c("Control", "One-at-a-time", "Leave-one-out", "Shapley")
kableExtra::kable_styling(
  knitr::kable(
    diagnostic_display,
    format = "latex",
    booktabs = TRUE,
    digits = 8,
    caption = "Three control-level diagnostics in the deterministic correlated-control example."
  ),
  latex_options = "HOLD_position"
)
```

The one-at-a-time and LOO columns answer different sensitivity questions, so neither is an additive allocation. The Shapley column is the only one constructed to allocate the joint shift across controls.

## Order dependence

```{r order-dependence, include=FALSE}
order_123 <- sequential_contributions(y, d, W, Z, c("Z1", "Z2", "Z3"))
order_321 <- sequential_contributions(y, d, W, Z, c("Z3", "Z2", "Z1"))
order_comparison <- merge(
  order_123[, c("control", "contribution")],
  order_321[, c("control", "contribution")],
  by = "control",
  suffixes = c(" (Z1 -> Z2 -> Z3)", " (Z3 -> Z2 -> Z1)"),
  sort = FALSE
)
order_comparison <- order_comparison[match(colnames(Z), order_comparison$control), ]
order_sums <- c(sum(order_123$contribution), sum(order_321$contribution))

all_orders <- .all_permutations(colnames(Z))
all_order_sums <- vapply(seq_len(nrow(all_orders)), function(index) {
  sum(sequential_contributions(y, d, W, Z, all_orders[index, ])$contribution)
}, numeric(1))
```

```{r order-dependence-table}
kableExtra::kable_styling(
  knitr::kable(
    order_comparison,
    format = "latex",
    booktabs = TRUE,
    digits = 8,
    caption = "Sequential contributions under two valid inclusion orders."
  ),
  latex_options = "HOLD_position"
)
```

The two paths both sum to the same joint shift (`r format(joint_shift, digits = 12)`), but the entries assigned to individual controls differ. This is order dependence, not a numerical error.

# Reproducibility checks

```{r validation, include=FALSE}
tolerance <- 1e-10
shapley_error <- abs(sum(shapley_results$shapley) - joint_shift)
sequential_error <- max(abs(all_order_sums - joint_shift))
order_difference <- max(abs(order_comparison[[2L]] - order_comparison[[3L]]))
minimum_control_correlation <- min(abs(
  stats::cor(Z)[upper.tri(stats::cor(Z))]
))

permuted_Z <- Z[, c("Z3", "Z1", "Z2"), drop = FALSE]
permuted_shapley <- shapley_allocation(y, d, W, permuted_Z)
column_order_error <- max(abs(
  shapley_results$shapley[match(colnames(Z), shapley_results$control)] -
    permuted_shapley$shapley[match(colnames(Z), permuted_shapley$control)]
))

rename_map <- c(Z1 = "Alpha", Z2 = "Beta", Z3 = "Gamma")
renamed_Z <- Z
colnames(renamed_Z) <- unname(rename_map[colnames(Z)])
renamed_shapley <- shapley_allocation(y, d, W, renamed_Z)
original_names <- names(rename_map)[
  match(renamed_shapley$control, unname(rename_map))
]
relabeling_error <- max(abs(
  shapley_results$shapley[match(original_names, shapley_results$control)] -
    renamed_shapley$shapley
))

validation_table <- data.frame(
  check = c(
    "Shapley contributions sum to the joint shift",
    "All six sequential paths telescope to the joint shift",
    "Shapley is invariant to candidate-column order",
    "Shapley is invariant to bijective variable relabeling",
    "Sequential control contributions differ across displayed orders",
    "One-at-a-time and LOO diagnostics differ for at least one control",
    "Candidate controls are substantively correlated"
  ),
  metric = c(
    shapley_error,
    sequential_error,
    column_order_error,
    relabeling_error,
    order_difference,
    max(abs(diagnostic_table$one_at_a_time - diagnostic_table$leave_one_out)),
    minimum_control_correlation
  ),
  criterion = c(
    "error < 1e-10", "error < 1e-10", "error < 1e-10",
    "error < 1e-10", "difference > 1e-4", "difference > 1e-4",
    "minimum |correlation| > 0.20"
  ),
  stringsAsFactors = FALSE
)
validation_table$result <- ifelse(
  c(
    shapley_error < tolerance,
    sequential_error < tolerance,
    column_order_error < tolerance,
    relabeling_error < tolerance,
    order_difference > 1e-4,
    validation_table$metric[6L] > 1e-4,
    minimum_control_correlation > 0.20
  ),
  "PASS",
  "FAIL"
)
stopifnot(all(validation_table$result == "PASS"))
```

```{r validation-table}
validation_display <- validation_table
validation_display$metric <- format(validation_display$metric, scientific = TRUE, digits = 4)
kableExtra::kable_styling(
  knitr::kable(
    validation_display,
    format = "latex",
    booktabs = TRUE,
    caption = "Reproducibility checks for the allocation diagnostics."
  ),
  latex_options = c("HOLD_position", "scale_down")
)
```

The column-order check changes the input order and realigns estimates by control name. The separate relabeling check assigns a deterministic one-to-one set of new names, maps the results back, and verifies that labels do not affect the allocation.

# Reporting guidance and limitations

For the paper's linear TSCS/CET use case, report the endpoint shift and LOO increments as transparent conditional sensitivity checks. Report one-at-a-time increments only when the short-model comparison is substantively useful. Reserve Shapley values for an online appendix or replication package: they are a coherent symmetric allocation of a projection shift, but they do not determine a unique causal contribution of correlated controls.

The functions in `scripts/shift_allocation_correlated_controls.R` require numeric finite inputs, a common complete sample, and full rank for every requested nested model. They enumerate all $q!$ orders and stop by default above nine controls, which is appropriate for a small, prespecified set of candidate controls but not automatically for high-dimensional specifications. They do not cover weighted least squares, non-nested models, regularized estimators, missing-data-induced sample changes, or causal identification of the CET.
